---
title: "Selecting_species_for_analysis"
author: "Lana Bolin"
date: "`r Sys.Date()`"
output: html_document
---


# Prep

```{r, message=F}
# Clear environment
rm(list = ls())

# Load in libraries
library(dplyr)
library(ggplot2)
library(ggsignif)
library(cowplot)
library(magrittr)
library(lme4)
library(car)
library(emmeans)
library(lmerTest)
library(cowplot)  # paneled figures
library(tidyr)   # For making soil moisture data long form
library(tidyverse)
library(DHARMa)
library(MuMIn) # for AICc function
library(sjPlot) # for plotting non-interaction marginal effect
library(scales) # to emmulate ggplot color palate in interactions( #)

# Load the dataset. Each row is a quadrat.
setwd("~/GitHub/SEV_Species_Interactions_Stability")
#data.biomass <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/plant_species_stability_traits.csv", stringsAsFactors = T)




```

Load and clean species data from EDI
```{r}
# Package ID: knb-lter-sev.318.1 Cataloging System:https://pasta.edirepository.org.
# Data set title: Sevilleta Long Term Ecological Research Program Plant Species List.
# Data set creator:  Stephanie Baker - University of New Mexico 
# Data set creator:  Lauren Baur - University of New Mexico 
# Data set creator:  Esteban Muldavin - University of New Mexico 
# Data set creator:  Jennifer Rudgers - University of New Mexico 
# Contact:  Kristofer Hall -  University of New Mexico  - khall001@unm.edu
# Stylesheet v2.11 for metadata conversion into program: John H. Porter, Univ. Virginia, jporter@virginia.edu 

inUrl1  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-sev/318/1/fdfe27e194a020d066acd8f11aa08dae" 
infile1 <- tempfile()
try(download.file(inUrl1,infile1,method="curl"))
if (is.na(file.size(infile1))) download.file(inUrl1,infile1,method="auto")

                   
 dt1 <-read.csv(infile1,header=F 
          ,skip=1
            ,sep=","  
        , col.names=c(
                    "kartez",     
                    "family",     
                    "genus",     
                    "sp_epithet",     
                    "subspecies_variety",     
                    "common_name",     
                    "Native",     
                    "PhotoPath",     
                    "LifeHistory",     
                    "FunctionalGroup"    ), check.names=TRUE)
               
unlink(infile1)
		    
# Fix any interval or ratio columns mistakenly read in as nominal and nominal columns read as numeric or dates read as strings
                
if (class(dt1$kartez)!="factor") dt1$kartez<- as.factor(dt1$kartez)
if (class(dt1$family)!="factor") dt1$family<- as.factor(dt1$family)
if (class(dt1$genus)!="factor") dt1$genus<- as.factor(dt1$genus)
if (class(dt1$sp_epithet)!="factor") dt1$sp_epithet<- as.factor(dt1$sp_epithet)
if (class(dt1$subspecies_variety)!="factor") dt1$subspecies_variety<- as.factor(dt1$subspecies_variety)
if (class(dt1$common_name)!="factor") dt1$common_name<- as.factor(dt1$common_name)
if (class(dt1$Native)!="factor") dt1$Native<- as.factor(dt1$Native)
if (class(dt1$PhotoPath)!="factor") dt1$PhotoPath<- as.factor(dt1$PhotoPath)
if (class(dt1$LifeHistory)!="factor") dt1$LifeHistory<- as.factor(dt1$LifeHistory)
if (class(dt1$FunctionalGroup)!="factor") dt1$FunctionalGroup<- as.factor(dt1$FunctionalGroup)
                
# Convert Missing Values to NA for non-dates
                
dt1$sp_epithet <- as.factor(ifelse((trimws(as.character(dt1$sp_epithet))==trimws("NA")),NA,as.character(dt1$sp_epithet)))
dt1$Native <- as.factor(ifelse((trimws(as.character(dt1$Native))==trimws("NA")),NA,as.character(dt1$Native)))
dt1$PhotoPath <- as.factor(ifelse((trimws(as.character(dt1$PhotoPath))==trimws("NA")),NA,as.character(dt1$PhotoPath)))
dt1$LifeHistory <- as.factor(ifelse((trimws(as.character(dt1$LifeHistory))==trimws("NA")),NA,as.character(dt1$LifeHistory)))
dt1$FunctionalGroup <- as.factor(ifelse((trimws(as.character(dt1$FunctionalGroup))==trimws("NA")),NA,as.character(dt1$FunctionalGroup)))


# Here is the structure of the input data frame:
str(dt1)                            
attach(dt1)                            
# The analyses below are basic descriptions of the variables. After testing, they should be replaced.                 

summary(kartez)
summary(family)
summary(genus)
summary(sp_epithet)
summary(subspecies_variety)
summary(common_name)
summary(Native)
summary(PhotoPath)
summary(LifeHistory)
summary(FunctionalGroup) 
                # Get more details on character variables
                 
summary(as.factor(dt1$kartez)) 
summary(as.factor(dt1$family)) 
summary(as.factor(dt1$genus)) 
summary(as.factor(dt1$sp_epithet)) 
summary(as.factor(dt1$subspecies_variety)) 
summary(as.factor(dt1$common_name)) 
summary(as.factor(dt1$Native)) 
summary(as.factor(dt1$PhotoPath)) 
summary(as.factor(dt1$LifeHistory)) 
summary(as.factor(dt1$FunctionalGroup))
detach(dt1)               
        
data.species <- dt1
```

```{r}
options(HTTPUserAgent="EDI_CodeGen")
          
inUrl1  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-sev/331/3/cb45bdfc2edb50ef242d29f16055ab5b" 
infile1 <- tempfile()
try(download.file(inUrl1,infile1,method="curl",extra=paste0(' -A "',getOption("HTTPUserAgent"),'"')))
if (is.na(file.size(infile1))) download.file(inUrl1,infile1,method="auto")

mass <-read.csv(infile1,header=F 
                , stringsAsFactors = T
                ,skip=1
                ,sep=","  
                ,quot='"' 
                , col.names=c(
                    "site",     
                    "year",     
                    "season",     
                    "date",     
                    "web",     
                    "transect",     
                    "block",     
                    "plot",     
                    "subplot",     
                    "quad",     
                    "treatment",     
                    "kartez",     
                    "genus",     
                    "sp.epithet",     
                    "family",     
                    "LifeHistory",     
                    "PhotoPath",     
                    "FunctionalGroup",     
                    "cover",     
                    "volume",     
                    "biomass.BM",     
                    "biomass.BIM",     
                    "SiteCluster",     
                    "MetStation",     
                    "season.precip",     
                    "GDD",     
                    "SPEI.comp"    ), check.names=TRUE)
               
unlink(infile1)

# Set plot theme
theme_set(theme_bw() +
            theme(
              plot.background = element_blank()
              ,panel.grid.major = element_blank()
              ,panel.grid.minor = element_blank()
              ,panel.background = element_blank()
              ,axis.text.x  = element_text(angle=90, vjust=0.5, size=8)
            ))

mass$web <- as.factor(mass$web)

### recode treatment into three columns for factorial 2 x 2 x 2 design (NOTE: BELOW I'LL NEED TO SUBSET TO treatment == C, N, P, T, TN, PN, TP, TPN)
# note 'T' requires quotes so it is not read as TRUE
mass$warm <- recode_factor(mass$treatment, C="Ambient_T",N="Ambient_T",P="Ambient_T",'T'="Warmed",TN="Warmed",PN="Ambient_T",TP="Warmed",TPN="Warmed")

mass$ppt <- recode_factor(mass$treatment, C="Ambient_ppt",N="Ambient_ppt",P="Extra_winter_ppt",'T'="Ambient_ppt",TN="Ambient_ppt",PN="Extra_winter_ppt",TP="Extra_winter_ppt",TPN="Extra_winter_ppt")

mass$nitrogen <- recode_factor(mass$treatment, C="Ambient_N",N="N_addition",P="Ambient_N",'T'="Ambient_N",TN="N_addition",PN="N_addition",TP="Ambient_N",TPN="N_addition")

data.biomass <- mass
```

Some data cleaning
```{r}
# Make unique quad_ID
data.biomass$quad_ID <- as.factor(
      paste(
        data.biomass$site,
        data.biomass$web,
        data.biomass$transect,
        data.biomass$treatment,
        data.biomass$block,
        data.biomass$plot,
        data.biomass$subplot,
        data.biomass$quad,
        sep = "_"))

# make a time vector by combining year and season
data.biomass$seas <- as.numeric(as.character(recode_factor(data.biomass$season, fall = "0.5", spring = "0")))
data.biomass$year_seas <- data.biomass$year + data.biomass$seas
data.biomass$time <- as.numeric(as.factor(data.biomass$year_seas))

# also make time point a factor
data.biomass$year_seas.f <- as.factor(data.biomass$year_seas)

# Only keep core sites
data.biomass.core <- data.biomass %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ")) 

length(unique(data.biomass.core$kartez))
```

Filter the quadrats to only include those with some minimum number of years of data (years_count column counts the number of years the species could have been observed because that quadrat was surveyed). I chose 10 years or more.
```{r}
QuadsToKeep <- data.biomass.core %>%
  group_by(quad_ID, year) %>%
  summarise(n = n()) %>%
  group_by(quad_ID) %>%
  summarise(n.years = n()) %>%
  filter(n.years >= 10)

data.biomass.core %<>%
  filter(quad_ID %in% QuadsToKeep$quad_ID)
```

Select species that have at least 5 years of non-zero biomass.
```{r}
# Count the number of years each species has non-zero biomass across the SEV. Filter to species with at least 5 years
SpeciesToKeep <- data.biomass.core %>%
  group_by(kartez, year) %>%
  summarise(n = n()) %>% # Get number of quads per species per year
  group_by(kartez) %>%
  summarise(n.years = n()) %>% # Get number of years that each species has biomass data
  filter(n.years >= 5) # Filter to species with 5 or more years of biomass data

data.biomass.core %<>%
  filter(kartez %in% SpeciesToKeep$kartez)

length(unique(data.biomass.core$kartez))
```

Combine with species data and write species list to csv
```{r}
data <- left_join(data.biomass.core, data.species, by = "kartez")

species.list <- data.biomass.core %>%
  select(kartez, genus, sp.epithet, LifeHistory, FunctionalGroup, PhotoPath) %>%
  filter(!kartez %in% c("LATR2", "YUGL", "YUBA", "PIED", "JUMO", "UNKNOWN", "EMPTY")) %>%
  unique(.)

nrow(species.list)
length(unique(species.list$kartez))

# Import old species list
old.species.list <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/species.list.csv", stringsAsFactors = T)

# Get list of species from the new species list that do not overlap with the old species list
species.list.nonoverlapping <- subset(species.list, !(kartez %in% old.species.list$kartez))
length(unique(species.list.nonoverlapping$kartez))

# Write CSV of non-overlapping species
write.csv(species.list.nonoverlapping, "species.list.nonoverlapping.csv", row.names = F)



# Write CSV of full longer list of species
write.csv(species.list, "species.list.longer.csv", row.names = F)



```





This code was for the plant_species_stability_traits.csv data, which is a different format than the data I want to use
```{r, eval=F}
data.biomass.core <- data.biomass %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ")) 

length(unique(data.biomass.core$kartez))

# Sum the biomass for each species across all sites and quadrats within each year. Columns 16:39 are the data for each year, so look across those columns. 24 total years.
yearsums <- data.biomass.core %>% group_by(kartez) %>%
  mutate_at(vars(16:39), ~replace_na(., 0)) %>%  # Change NAs to 0's for yearly biomass data
  summarise_each(funs(sum), X1999, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022)  # Get total biomass for each species in each year (across sites & quadrats)

# Count the number of years each species has non-zero biomass across the SEV. Add as column to yearsums
yearsums$n.nonzero.years <- rowSums(yearsums[, 2:25] != 0)

# Filter to years with at least 3 years producing biomass 
ThreePlus <- yearsums %>%
  filter(n.nonzero.years >=3) 

# Filter out species with fewer than 3 non-zero years
data.more3 <- data.biomass.core %>%
  filter(kartez %in% ThreePlus$kartez)

length(unique(data.more3$kartez))
```

Combine with species data - OLD
```{r, eval=F}
data <- left_join(data.more3, data.species, by = "kartez")

species.list <- data %>%
  select(kartez, genus, sp_epithet, LifeHistory, FunctionalGroup, PhotoPath) %>%
  unique(.)

write.csv(species.list, "species.list.csv", row.names = F)
```


















