---
title: "SEV_Sp_Intrxns_Kamren_Analyses_2"
author: "Lana Bolin"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
---

USE THIS SCRIPT. 

There is an other script called "SEV_Sp_Intrxns_Kamren_Analyses" that imports the wide-format data, but don't use that one. use this one.


# Parking downhill

* GO THROUGH WHOLE SCRIPT AND FIX STUFF FROM FILTERING TO FALL ONLY

* Try merging bee, mammal, and plant data. Plot plant CV vs. bee CV, using the mean CV per site (only 3 sites, so there will only be 3 data points).
* Ara is having a look at the brms models and trouble-shooting. He's going to try reducing the step size, playing with the priors, and replacing quad_ID with lat/long.

# To do

* Is there a way to estimate variability for a binary response? If so, do that for legumes (presence/absence). I can also try analyzing count data.
* Separate data into spring and fall (dry vs. monsoon season), and see if trends differ.



# Setup

## Load and clean data

```{r, message=F}
# Clear environment
rm(list = ls())

# Load libraries
libraries <- c("magrittr", "lme4", "car", "emmeans", "MuMIn", "scales", "sjPlot", "vegan", "ape", "phytools", "devtools", "ggbiplot", "tidyverse", "googlesheets4", "brms", "reshape2", "DHARMa", "ggeffects")
invisible(lapply(libraries, require, character.only = T))

setwd("~/GitHub/SEV_Species_Interactions_Stability")

# Set plot theme
theme_set(theme_bw() +
            theme(
              plot.background = element_blank()
              ,panel.grid.major = element_blank()
              ,panel.grid.minor = element_blank()
              ,panel.background = element_blank()
              ,axis.text.x  = element_text(vjust=0.5, size=10)
            ))

### Load the datasets

# Interaction data. Each row is a quadrat.
data.intrxns <- read_sheet("https://docs.google.com/spreadsheets/d/1hz3HQ7cWakZdqdvxTCEz2sbjjtqssulhvIh4sBBYEI8/edit?gid=0#gid=0", "Data")

# Quad data. Has quad GPS coordinates for Bayesian analysis
quad.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/NPP_quads_latlong_edited.csv", stringsAsFactors = T) 

# Read in species data to get native/invasive status
data.species <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/SevilletaSpeciesList2019.csv", stringsAsFactors = T) 

# Import data that Jenn sent (she was able to run my code - see email from 4/1/2025). Wide format, columns for each year.
mass_species_cast <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/mass_species_cast.csv")

# Read in to get plant families
family.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/plant_species_stability_traits_2024.csv", stringsAsFactors = T)
```

### Clean data

Prep each dataset
```{r}
### mass_species_cast

# Filter species_CV data to only include "C" treatment
mass_species_cast %<>% filter(treatment == "C")

# Fall only
mass_species_cast %<>% filter(season == "fall")


### data.species

# Fix typo in species data
data.species$Native[data.species$Native == "Y"] <- "y"

# Select columns I need
data.species %<>% dplyr::select(kartez, Native)


### data.intrxns

# Rename columns from species interactions data sheet
data.intrxns %<>% 
  rename("Legume" = "Legume? (Y/N)",
         "LegumeSource" = "Source: Legume",
         "Pollination" = "Pollination (W/A/WA)",
         "PollinationGS" = "Pollination: genus or species? (G/S)",
         "PollinationSource" = "Source: Pollination",
         "Dispersal" = "Dispersal (W/A/WA)",
         "DispersalGS" = "Dispersal: genus or species? (G/S)",
         "DispersalCheekStomach" = "Dispersal: cheek, stomach, or both? (only for if source is Hope & Parmenter)",
         "DispersalSource" = "Source: Dispersal",
         "Outcrossing" = "Outcrossing (O/N)",
         "OutcrossingGS" = "Outcrossing: genus or species? (G/S)",
         "OutcrossingSource" = "Source: Outcrossing",
         "Hybridizing" = "Hybridizing (H/N)",
         "HybridizingGS" = "Hybridizing: genus or species? (G/S)",
         "HybrydizingSource" = "Source: Hybridizing",
         "NotesInteraction" = "Notes") %>%
  dplyr::select(-c(genus, species))

# Fix LifeHistory 
data.intrxns %<>% mutate(LifeHistory = ifelse(LifeHistory %in% c("a", "annual"), "annual",
                                             ifelse(LifeHistory %in% c("p", "perennial"), "perennial", NA)))

# Create new columns for pollination and dispersal that change "WA" to "W". The logic is, if the species can be wind pollinated, it doesn't rely on other species.
data.intrxns %<>%
  mutate(Pollination.new = as.factor(ifelse(Pollination == "A", "A", "W")),
         Dispersal.new = as.factor(ifelse(Dispersal == "A", "A", "W")))

# Change all character vectors to factors
data.intrxns %<>% mutate_if(is.character, as.factor)

# Fix FunctionalGroup levels
data.intrxns %<>%
  mutate(FunctionalGroup = ifelse(FunctionalGroup %in% c("f", "forb"), "forb",
                                  ifelse(FunctionalGroup %in% c("g", "grass"), "grass",
                                         ifelse(FunctionalGroup %in% c("s", "shrub"), "shrub", NA))))


### quad.data

quad.data$quad_ID <- as.factor(
      paste(
        quad.data$site,
        quad.data$web,
        quad.data$transect,
        quad.data$treatment,
        quad.data$block,
        quad.data$plot,
        quad.data$subplot,
        quad.data$quad,
        sep = "_"))

quad.data$web <- as.factor(quad.data$web)
quad.data$block <- as.factor(quad.data$web)
quad.data$quad <- as.factor(quad.data$quad)


# family.data

family.data %<>%
  dplyr::select(kartez, family) %>%
  group_by(kartez) %>%
  distinct()
```

Merge datasets
```{r}
# Merge species interaction data, native/invasive status, and wide data
data.mass <- left_join(data.intrxns, data.species, by = "kartez")
data.mass <- left_join(data.mass, mass_species_cast, by = "kartez")

# Change characters to factors
data.mass %<>% mutate_if(is.character, as.factor)
data.mass$block <- as.factor(data.mass$block)
data.mass$web <- as.factor(data.mass$web)
data.mass$block <- as.factor(data.mass$web)
data.mass$quad <- as.factor(data.mass$quad)

# Make sure levels are what I'm expecting
levels(data.mass$Pollination)
levels(data.mass$Dispersal) 

# NOMI is the only species with "(likely W)A" dispersal level. Let's call it WA.
data.mass %>% filter(Dispersal == "(likely W)A") %>% group_by(kartez) %>% summarise(n = n())
data.mass %<>% mutate(Dispersal = fct_recode(Dispersal, "WA" = "(likely W)A"))

data.mass <- as.data.frame(data.mass)

data.mass$X <- NULL

# Filter out unidentified species (that were only identified to genus) "Sporobolus_NA", "Opuntia_NA", "Astragalus_NA", and "Aristida_NA". These are kartez: ASTRA, ARIST, OPUNT, SPORO
data.mass %<>% filter(!kartez %in% c("SPORO", "OPUNT", "ASTRA", "ARIST")) 

nrow(data.mass)

# Merge quad coordinates with data
data <- left_join(data.mass, quad.data, by = c("site", "web", "transect", "block", "plot", "subplot", "quad", "treatment", "quad_ID"))

nrow(data)
```



Start everybody at the start date of the shortest site because longer time series –> pick up more variability (need same time series length).
```{r}
# Find the earliest year when there is data for all of the sites. I'm going to filter out the earlier years after filtering to quadrats with at least 10 years of data and checking that some species don't have mostly zero abundance in one year. Maybe make a dataset starting in 2004 (7 sites), and one starting in 2006 (9 sites), 2008 (10 sites), 2011 (11 sites), and 2012 (13 sites).

# I may also want to calculate CV on dataset restricted to these years... Probably I guess? Also for SEV plant traits analysis!

data %>%
  group_by(site) %>%
  summarise(X1999.yrs = sum(!is.na(X1999)),
            X2000.yrs = sum(!is.na(X2000)),
            X2001.yrs = sum(!is.na(X2001)),
            X2002.yrs = sum(!is.na(X2002)),
            X2003.yrs = sum(!is.na(X2003)),
            X2004.yrs = sum(!is.na(X2004)),
            X2005.yrs = sum(!is.na(X2005)),
            X2006.yrs = sum(!is.na(X2006)),
            X2007.yrs = sum(!is.na(X2007)),
            X2008.yrs = sum(!is.na(X2008)),
            X2009.yrs = sum(!is.na(X2009)),
            X2010.yrs = sum(!is.na(X2010)),
            X2011.yrs = sum(!is.na(X2011)),
            X2012.yrs = sum(!is.na(X2012)),
            X2013.yrs = sum(!is.na(X2013)),
            X2014.yrs = sum(!is.na(X2014)),
            X2015.yrs = sum(!is.na(X2015)),
            X2016.yrs = sum(!is.na(X2016)),
            X2017.yrs = sum(!is.na(X2017)),
            X2018.yrs = sum(!is.na(X2018)),
            X2019.yrs = sum(!is.na(X2019)),
            X2020.yrs = sum(!is.na(X2020)),
            X2021.yrs = sum(!is.na(X2021)),
            X2022.yrs = sum(!is.na(X2022)),
            X2023.yrs = sum(!is.na(X2023))) %>%
  print(n=Inf)

data.2004 <- data %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ", "fertilizer", "mixed_grass", "mixed_shrub")) %>%
  dplyr::select(-c("X1999", "X2000", "X2001", "X2002", "X2003")) # Remove years that aren' included in this dataset, so I can calculate species_CV and species_mu across these years (everybody should have same time series)

data.2012 <- data %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ", "fertilizer", "mixed_grass", "mixed_shrub", "MRME", "warming", "NutNet", "tower_east", "tower_west", "EDGE_blue", "EDGE_black")) %>% 
  dplyr::select(-c("X1999", "X2000", "X2001", "X2002", "X2003", "X2004", "X2005", "X2006", "X2007", "X2008", "X2009", "X2010", "X2011")) # Remove years that aren' included in this dataset, so I can calculate species_CV and species_mu across these years (everybody should have same time series)

# Using the 2012 data, this leaves 158 species. Two of Kamren's species are no longer in the dataset. That's pretty good!
length(unique(data.2012$kartez))
length(unique(data$kartez))

# These are the 2 species not present in full data:
setdiff(data$kartez, data.2012$kartez)
```


Calculate species_CV for each species in each quad, in each dataset. I'm going to focus on 2004 and 2012 for now.
```{r}
# List of dfs
dfs.list <- list(data.2004, data.2012)

# List of df names
dfs.list.names <- list("data.2004", "data.2012")

# List of end column numbers (final year of dataset)
end.cols <- list(55, 47)

final.dfs <- list()

for (i in 1:length(dfs.list)) {
  
  temp.df <- dfs.list[[i]]
  temp.df.name <- dfs.list.names[[i]]
  endcol <- end.cols[[i]]
  
  final.dfs[[temp.df.name]] <- temp.df %>%
    mutate(biomass.sum = apply(.[(36:endcol)], 1, sum, na.rm=T),
           species_mu = apply(.[(36:endcol)], 1, mean, na.rm=T),
           species_sd = apply(.[(36:endcol)], 1, sd, na.rm=T),
           species_CV = species_sd/species_mu) %>%
    filter(biomass.sum > 0) # Remove all species that were never observed in a quad
  
}

data.2004 <- final.dfs$data.2004  # 20 years of time series, 7 sites (fewer rows of data)
data.2012 <- final.dfs$data.2012  # 12 years of time series, 14 sites (more rows of data)

# We lose 1 species and gain 8,312 rows of data (quad x species combos) by using 2012 data.
length(unique(data.2004$kartez))
length(unique(data.2012$kartez))

nrow(data.2004)
nrow(data.2012)

nrow(data.2012) - nrow(data.2004)

# Write 2012 data to csv for Kamren to play with
write.csv(data.2012, "~/GitHub/SEV_Species_Interactions_Stability/data/data.2012.csv", row.names = F)
```

Filter out species with less than 3 years of non-zero biomass data.
```{r}
# Sum the biomass for each species across all sites and quadrats within each year. Columns 16:39 are the data for each year, so look across those columns. 24 total years.
yearsums <- data.2012 %>% 
  group_by(kartez) %>%
  summarise(across(c(X2012, X2013, X2014, X2015, X2016, X2017, X2018, X2019, X2020, X2021, X2022, X2023), sum, na.rm=T)) # Get total biomass for each species in each year (across sites & quadrats)

# Count the number of years each species has non-zero biomass across the SEV. Add as column to yearsums
yearsums$n.nonzero.years <- rowSums(yearsums[, 2:13] != 0)

# Filter to species with at least 3 years producing biomass.
ThreePlus <- yearsums %>%
  filter(n.nonzero.years >=3) 

data.2012 %<>%
  filter(kartez %in% ThreePlus$kartez)

# This step filtered out 27 species (from 156 to 129)
length(unique(data.2012$kartez))
```

### Data summaries

Overall summaries
```{r}
# Number of species
length(unique(data.2012$kartez))

# Number of sites, and quads per site. 4 ecosystems (the 4 core sites) and 2 ecotones (grass-grass and grass-shrub). 
# Grass-grass ecotone: mixed grass, warming, fertilizer, nutnet, plus one of the two tower sites. 
# Grass-shrub ecotone: mixed_grass
# Can look at EDI metadata (geography section) to see which sites are in which ecosystems
data.2012 %>% group_by(site) %>% summarise(n.quads = length(unique(quad_ID)))

# Number of quads
length(unique(data.2012$quad_ID))



# Number of species with data for each species interaction
data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(Pollination.new) %>%
  summarise(n = n())

data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(Dispersal.new) %>%
  summarise(n = n())

# Number of annuals vs. perennials
data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(LifeHistory) %>%
  summarise(n = n())

# Breakdown of pollination strategy by annual/perennial 
data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(LifeHistory, Pollination.new) %>%
  summarise(n = n())

  # % of annuals that are animal-pollinated 
  22/(22+14)*100
  
  # % of perennials that are animal-pollinated
  65/(65+27)*100
  
# Breakdown of dispersal strategy by annual/perennial
data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(LifeHistory, Dispersal.new) %>%
  summarise(n = n())

  # % of annuals that are animal-dispersed 
  30/(30+5)*100
  
  # % of perennials that are animal-dispersed 
  67/(67+25)*100

# Breakdown of dispersal strategy by grass/non-grass
data.intrxns %>% 
  filter(kartez %in% data.2012$kartez) %>%
  group_by(FunctionalGroup, Dispersal.new) %>%
  summarise(n = n())
  
  # % of grasses that are animal-dispersed
  20/(20+7)*100

  # % of non-grasses that are animal-dispersed
  (64+13)/(64+13+16+7)*100

# How many species had genus-level pollination data?
data.2012 %>%
  group_by(kartez) %>%
  filter(row_number() == 1) %>%
  group_by(PollinationGS) %>%
  summarise(n = n())

# How many species had genus-level dispersal data?
data.2012 %>%
  group_by(kartez) %>%
  filter(row_number() == 1) %>%
  group_by(DispersalGS) %>%
  summarise(n = n())
```

How many families are in the dataset?
```{r}
data.2012 <- left_join(data.2012, family.data, by = "kartez")

families <- data.2012 %>%
  group_by(family) %>% 
  dplyr::summarise(n = n()) 

# Number of families in dataset
nrow(families)

# Families in dataset
data.2012 %>%
  group_by(family) %>% 
  dplyr::summarise(n = n())
```

How many species are in each category?
```{r}
### Pollination

# Original categories
table(data.intrxns$Pollination)
# With WA re-leveled as W
table(data.intrxns$Pollination.new)


### Dispersal

# Original categories
table(data.intrxns$Dispersal)
# With WA re-leveled as W
table(data.intrxns$Dispersal.new)


### Life History
table(data.intrxns$LifeHistory)

### Native

# Take one row of data per species
species.info <- data %>% 
  group_by(kartez) %>%
  filter(row_number()==1)

table(species.info$Native)

# What are the 4 non-native species?
species.info %>% 
  filter(Native == "n") %>%
  filter(row_number()==1) %>%
  dplyr::select(kartez)
```




Look at data
```{r}
# Pollination 
ggplot(data.2012, aes(x = Pollination.new, y = species_CV)) +
  geom_violin()

# Dispersal
ggplot(data.2012, aes(x = Dispersal.new, y = species_CV)) +
  geom_violin()
```



How much of the total plant biomass in the system is represented by the focal set?  

Make equivalent dataset with all species in those same quads/years/sites/season
```{r}
# total biomass of all species: 2012-2023, 14 sites, fall only
options(HTTPUserAgent="EDI_CodeGen")
          
inUrl1  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-sev/331/3/cb45bdfc2edb50ef242d29f16055ab5b" 
infile1 <- tempfile()
try(download.file(inUrl1,infile1,method="curl",extra=paste0(' -A "',getOption("HTTPUserAgent"),'"')))
if (is.na(file.size(infile1))) download.file(inUrl1,infile1,method="auto")

mass <-read.csv(infile1,header=F 
                , stringsAsFactors = T
                ,skip=1
                ,sep=","  
                ,quot='"' 
                , col.names=c(
                    "site",     
                    "year",     
                    "season",     
                    "date",     
                    "web",     
                    "transect",     
                    "block",     
                    "plot",     
                    "subplot",     
                    "quad",     
                    "treatment",     
                    "kartez",     
                    "genus",     
                    "sp.epithet",     
                    "family",     
                    "LifeHistory",     
                    "PhotoPath",     
                    "FunctionalGroup",     
                    "cover",     
                    "volume",     
                    "biomass.BM",     
                    "biomass.BIM",     
                    "SiteCluster",     
                    "MetStation",     
                    "season.precip",     
                    "GDD",     
                    "SPEI.comp"    ), check.names=TRUE)
               
unlink(infile1)

# Make unique quad_ID
mass$quad_ID <- as.factor(
      paste(
        mass$site,
        mass$web,
        mass$transect,
        mass$treatment,
        mass$block,
        mass$plot,
        mass$subplot,
        mass$quad,
        sep = "_"))

# Filter to the same years, quads, season, and sites
mass.filt <- mass %>% 
  filter(year %in% 2012:2023) %>%
  filter(quad_ID %in% data.2012$quad_ID) %>%
  filter(season == "fall") %>%
  filter(site %in% data.2012$site)

# Focal set total biomass: 2012-2023, 14 sites, fall only, filtered quads
focal.sum <- sum(data.2012$biomass.sum)

# Total biomass of all plants in the filtered quads in 2012-2023, 14 sites, fall only
total.sum <- sum(mass.filt$biomass.BM)

# % of total biomass represented by our focal set
focal.sum / total.sum * 100
  
```





# Models (2012-2023)

## Main model

```{r, message=F}
### Full model
mod.all <- lmer(species_CV ~ Pollination.new + Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(kartez != "CHER2")) # CHER2 only has family-level data 
Anova(mod.all, type = 2)
plot(mod.all)

### Separate model for each predictor
mod.pol <- lmer(species_CV ~ Pollination.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(kartez != "CHER2")) # CHER2 only has family-level data 
mod.disp <- lmer(species_CV ~ Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012)

Anova(mod.pol, type = 2)
Anova(mod.disp, type = 2)

### Effect sizes (% increase for animal-pollinated or -dispersed)

# Pollination
pol.emm <- summary(emmeans(mod.pol, pairwise ~ Pollination.new))
(pol.emm$emmeans[1, 2] - pol.emm$emmeans[2, 2]) / pol.emm$emmeans[2, 2] * 100

# Dispersal
disp.emm <- summary(emmeans(mod.disp, pairwise ~ Dispersal.new))
(disp.emm$emmeans[1, 2] - disp.emm$emmeans[2, 2]) / disp.emm$emmeans[2, 2] * 100


### Plot marginal effect of each predictor

## Pollination

# Get predictions
pred_pol <- ggpredict(mod.pol, terms = "Pollination.new")

# Plot
Pol.new.plot <- ggplot(pred_pol, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.05) +
  scale_x_discrete(
    limits = c("A", "W"),
    labels = c("Animal", "Wind")
  ) +
  scale_color_manual(
    values = c("A" = "red", "W" = "blue"),
    labels = c("A" = "Animal", "W" = "Wind")
  ) +
  labs(
    y = "Temporal Variability (Biomass CV)",
    x = "Pollination Strategy",
    color = "Pollination"
  ) +
  ylim(c(2.15, 3)) +
  theme(legend.position = "none")
Pol.new.plot


## Dispersal

# Get predictions
pred_disp <- ggpredict(mod.disp, terms = "Dispersal.new")

# Plot
Disp.new.plot <- ggplot(pred_disp, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.05) +
  scale_x_discrete(
    limits = c("A", "W"),
    labels = c("Animal", "Wind")
  ) +
  scale_color_manual(
    values = c("A" = "darkgreen", "W" = "orange"),
    labels = c("A" = "Animal", "W" = "Wind")
  ) +
  labs(
    y = "Temporal Variability (Biomass CV)",
    x = "Dispersal Strategy",
    color = "Dispersal"
  ) +
  ylim(c(2.15, 3)) +
  theme(legend.position = "none")
Disp.new.plot


ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Pol.new.plot.jpg", Pol.new.plot, dpi = 600, width = 2200, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Disp.new.plot.jpg", Disp.new.plot, dpi = 600, width = 2200, height = 2500, units = "px")
```


## Only including species for which we have species-level info

We lose too many species - leave this out of the MS but keep it in the code in case a reviewer asks for it.
```{r}
data.2012.pol.s <- data.2012 %>%
  filter(PollinationGS == "S")

# 63 species with species-level pollination data
length(unique(data.2012.pol.s$kartez))

data.2012.disp.s <- data.2012 %>%
  filter(DispersalGS == "S")

# 52 species with species-level pollination data
length(unique(data.2012.disp.s$kartez))

### Fit models
mod.pol.s <- lmer(species_CV ~ Pollination.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012.pol.s) 
mod.disp.s <- lmer(species_CV ~ Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012.disp.s)

Anova(mod.pol.s, type = 2)
Anova(mod.disp.s, type = 2)

#plot_model(mod.pol.s, type = "pred", terms = "Pollination.new", color = "black")
#plot_model(mod.disp.s, type = "pred", terms = "Dispersal.new", color = "black")
```


## Fit a model that includes the Pollination x Dispersal interaction

```{r}
mod.pol.disp.int <- lmer(species_CV ~ Pollination.new * Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012)

Anova(mod.pol.disp.int, type = 2)

plot.polxdisp <- plot_model(mod.pol.disp.int, type = "pred", terms = c("Pollination.new", "Dispersal.new"), dodge = 0.4) +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Pollination Strategy")
plot.polxdisp

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/plot.polxdisp.jpg", plot.polxdisp, dpi = 600, width = 3500, height = 2500, units = "px")
```


## Fit a model that includes the Pollination x LifeHistory interaction

```{r}
### Full model
mod.pol.lhist <- lmer(species_CV ~ Pollination.new * LifeHistory + X + Y + (1|kartez), na.action = na.omit, data = data.2012)

Anova(mod.pol.lhist, type = 2)

plot.polxLH <- plot_model(mod.pol.lhist, type = "pred", terms = c("Pollination.new", "LifeHistory"), dodge = 0.4) +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  scale_color_manual(
    values = c("annual" = "black", "perennial" = "darkgray"),  # or whichever colors you want
    labels = c("annual" = "Annual", "perennial" = "Perennial")
  ) +
  labs(y = "Temporal Variability (Biomass CV)", 
       x = "Pollination Strategy",
       color = "Life History") +
  theme(
  axis.title = element_text(size = 16),
  axis.text  = element_text(size = 14),
  axis.text.x = element_text(size = 14)
)
plot.polxLH

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/plot.polxLH.jpg", plot.polxLH, dpi = 600, width = 3100, height = 2300, units = "px")


# How many perennial wind-pollinated species are grasses? 21/27 species are grasses
data.2012 %>%
  filter(LifeHistory == "perennial" & Pollination.new == "W") %>%
  group_by(kartez, FunctionalGroup) %>%
  summarise(n=n()) %>%
  print(n=Inf)
```

## Try models that include species_mu
```{r, warning=F}
### Separate model for each predictor ###
mod.pol.mu <- lmer(species_CV ~ Pollination.new * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA"))
mod.disp.mu <- lmer(species_CV ~ Dispersal.new * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Dispersal != "WA"))

Anova(mod.pol.mu, type = 2)
Anova(mod.disp.mu, type = 2)

# Plot marginal effect of each predictor
Pol.mu.plot <- plot_model(mod.pol.mu, type = "pred", terms = "Pollination.new", color = "black") +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Pollination Strategy") +
  ylim(c(1.7, 2.9))

Pol.mu.plot


Disp.mu.plot <- plot_model(mod.disp.mu, type = "pred", terms = "Dispersal.new", color = "black")  +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Dispersal Strategy") +
  ylim(c(1.7, 2.9))

Disp.mu.plot

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Pol.mu.plot.jpg", Pol.mu.plot, dpi = 600, width = 2000, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Disp.mu.plot.jpg", Disp.mu.plot, dpi = 600, width = 2000, height = 2500, units = "px")


# Plot abundance against CV, and against sd
ggplot(data.2012, aes(x = species_mu, y = species_CV, color = Pollination.new)) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(data.2012, aes(x = species_mu, y = species_sd, color = Pollination.new)) +
  geom_point() +
  geom_smooth(method = "lm")
```


## Drop grasses from model, and see if pollination is still significant (p = 0.092)

```{r}
data.2012.nograss <- data.2012 %>%
  filter(!FunctionalGroup %in% c("g", "grass"))

### Separate model for each predictor
mod.pol.nograss <- lmer(species_CV ~ Pollination.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012.nograss %>% filter(kartez != "CHER2")) # Filter out the species that we only have family-level data for
mod.disp.nograss <- lmer(species_CV ~ Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2012.nograss)
mod.leg.nograss <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2012.nograss)

Anova(mod.pol.nograss, type = 2)
Anova(mod.disp.nograss, type = 2)
Anova(mod.leg.nograss, type = 2)

# % increase for marginally significant pollinator effect
pol.nograss.emm <- summary(emmeans(mod.pol.nograss, pairwise ~ Pollination.new))
(pol.nograss.emm$emmeans[1, 2] - pol.nograss.emm$emmeans[2, 2]) / pol.nograss.emm$emmeans[2, 2] * 100
```






# Outcrossing (with species as a random effect)

Data cleaning
```{r}
levels(data.2012$Outcrossing)

# For now, just use "O" and "N" categories 
data.2012.out <- data.2012 %>%
  filter(Outcrossing %in% c("O", "N"))

# Doing this removes 2 species, which is 579 rows of data
nrow(data.2012)
nrow(data.2012.out)

data.2012 %>%
  filter(Outcrossing %in% c("Mostly O", "O & M")) %>%
  distinct(kartez)

# How many species in each category?
out.cats <- data.intrxns %>%
  filter(Outcrossing %in% c("O", "N"))

table(out.cats$Outcrossing)[2:3]
```

Model
```{r}
# I need to change the order of "Outcrossing" levels for the plot
data.2012.out$Outcrossing <- factor(data.2012.out$Outcrossing, levels = c("O", "N"))

mod.out <- lmer(species_CV ~ Outcrossing + X + Y + (1|kartez), na.action = na.omit, data = data.2012.out)
Anova(mod.out, type = 2)
plot(mod.out)

outcross.plot <- plot_model(mod.out, type = "pred", terms = "Outcrossing", color = "black") +
  scale_x_discrete(limits = c("O", "N"),
                   labels = c("Obligately Outcrossing", "Not Obligately Outcrossing")) +
  labs(y = "CV of Species Biomass", x = "Outcrossing Method")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/outcross.plot.jpg", outcross.plot, dpi = 600, width = 3000, height = 2500, units = "px")
```







# Bee, bird, small mammal analyses - are bees and birds more unstable than mammals? 

Take each species at each location (“habitat”). Then calculate CV across locations. Exclude squirrels and chipmunks. Keep genera that start with “di” “ne” “on” (also try excluding these), “pe”, “pg”. There are multiple species of all of those. Exclude all other genera. Only use creosote and black grama sites ("habitat"). Analyze across all years of data (no need to restrict to Kamren’s dataset).  

But, I need bee and mammal data to have the same time frame as each other (so I can compare them).

```{r}
bee.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/BEE_DATA_knb-lter-sev.321.7/SEVBeeData2002-2019_revised2023-07-19.csv", stringsAsFactors = T)
mammal.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/SMALL_MAMMAL_DATA_knb-lter-sev.8.297978/SEV008_wide.csv", stringsAsFactors = T)


# For mammal data, only keep genera that start with “DI” “NE” “ON” (also try excluding these), “PE”, “PG”
mammal.data %<>%
  dplyr::select(-c("AMIN", "AMLE", "CHIN", "MUMU", "MUMU", "OTVA", "REME", "REMO", "REXX", "SIHI", 
"SPSP")) %>%
  #dplyr::select(-c("NEAL", "NEDO", "NEMI", "NEQU", "NEXX")) %>% # TEMP to check if removing wood rats (genus Neotoma) changes results. Wood rats had weird data for some reason - ask Jenn.
  filter(habitat %in% c("Black grama grassland", "Creosote shrubland")) %>%
  filter(season_fs == "fall")

```

Look at each species' species_CV (calculate for each species in each habitat/ecosystem), and see if they differ (we expect bees to be more sensitive). Make sure the data have the same number of observations per year (e.g., Jenn suggested taking max abundance of each species in each year to get a single annual estimate), and the same time series length.

## Bee data

Bee data: B = blue grama site, C = creosote site, G = black grama site
```{r}
levels(bee.data$ecosystem)
range(bee.data$year) # There are 18 years of bee data

# Filter to fall only (months 7-10)
bee.data %<>% filter(month %in% c(7:10))

### For each species, in each year, in each site, get the max abundance

# Melt into standard long format
bee.melt <- melt(bee.data, id.vars = c("month", "year", "start_date", "end_date", "complete_sampling_month", "complete_sampling_year", "ecosystem", "transect", "direction", "color"),
                 value.name = "count", variable.name = "species_code", na.rm = TRUE, factorsAsStrings = TRUE)

# There are 341 species
length(unique(bee.melt$species_code))

# Get max abundance in each species x year x ecosystem
bee.summary <- bee.melt %>%
  group_by(species_code, year, ecosystem) %>%
  summarise(max.abund = max(count))

# Cast data so there is a column for each year, and a row for each species x ecosystem
bee.wide <- dcast(bee.summary, species_code + ecosystem ~ year, sum, value.var = "max.abund", fill=0)

# Find how many years each bee species has non-zero abundance
bee.yearsums <- bee.wide %>%
  group_by(species_code) %>%
  summarise_each(funs(sum), 3:19)

# Count the number of years each species has non-zero biomass across the SEV. Add as column to yearsums
bee.yearsums$n.nonzero.years <- rowSums(bee.yearsums[, 2:18] != 0)

# Filter to years with <3 years producing biomass 
bee.less3 <- bee.yearsums %>%
  filter(n.nonzero.years < 3) 

# Filter out species with fewer than 3 non-zero years
bee.wide %<>%
  filter(!species_code %in% bee.less3$species_code)

# Calculate bee_CV as CV of max abundance over time, for each ecosystem
bee.wide %<>%
  mutate(sum = apply(.[(3:20)], 1, sum, na.rm=T),
           species_mu = apply(.[(3:20)], 1, mean, na.rm=T),
           species_sd = apply(.[(3:20)], 1, sd, na.rm=T),
           species_CV = species_sd/species_mu) %>%
  filter(!is.na(species_CV)) # Remove species x ecosystem combos where a species was never observed, because species_CV is NaN

```



## Mammal data

Data for each species represent "Number of animals per Sherman trap per night of the trapping bout for species".
```{r}
levels(mammal.data$ecosystem)
levels(mammal.data$habitat)
range(mammal.data$year)

# Jenn says to only use Creosote and Black Grama sites because Blue Grama site was collected for a short time.
mammal.data %<>%
  filter(habitat %in% c("Black grama grassland", "Creosote shrubland")) %>%
  dplyr::select(c(1, 5, 8:28)) #%>%
  #dplyr::select(c(1, 5, 8:23)) # TEMP for testing removing Neotoma genus


### For each species, in each year, in each site, get the max abundance

# Melt into standard long format
mammal.melt <- melt(mammal.data, id.vars = c("year", "habitat"),
                 value.name = "animals.trap.night", variable.name = "species_code", na.rm = TRUE, factorsAsStrings = TRUE)

# Filter to only include 2002-2019 (same years as bee data)
mammal.melt %<>%
  filter(year %in% c(2002:2019))

# There are 21 mammal species (16 with Neotoma excluded)
length(unique(mammal.melt$species_code))

# Get max abundance in each species x year x ecosystem
mammal.summary <- mammal.melt %>%
  group_by(species_code, year, habitat) %>%
  summarise(max.animals = max(animals.trap.night))

# Cast data so there is a column for each year, and a row for each species x habitat
mammal.wide <- dcast(mammal.summary, species_code + habitat ~ year, sum, value.var = "max.animals", fill=0)

# Find how many years each bee species has non-zero abundance
mammal.yearsums <- mammal.wide %>%
  group_by(species_code) %>%
  summarise_each(funs(sum), 3:19)

# Count the number of years each species has non-zero biomass across the SEV. Add as column to yearsums
mammal.yearsums$n.nonzero.years <- rowSums(mammal.yearsums[, 2:18] != 0)

# Filter to years with <5 years producing biomass 
mammal.less3 <- mammal.yearsums %>%
  filter(n.nonzero.years < 3) 

# Filter out species with fewer than 5 non-zero years
mammal.wide %<>%
  filter(!species_code %in% mammal.less3$species_code)

# Calculate bee_CV as CV of max abundance over time, for each ecosystem
mammal.wide %<>%
  mutate(sum = apply(.[(3:20)], 1, sum, na.rm=T),
           species_mu = apply(.[(3:20)], 1, mean, na.rm=T),
           species_sd = apply(.[(3:20)], 1, sd, na.rm=T),
           species_CV = species_sd/species_mu) %>%
  filter(!is.na(species_CV)) # Remove species x habitat combos where a species was never observed, because species_CV is NaN

# Rename "habitat" column to "ecosystem" to match bee data
mammal.wide %<>%
  rename("ecosystem" = "habitat")

# Relevel ecosytstem to match bee data
levels(bee.wide$ecosystem)
levels(mammal.wide$ecosystem)
levels(mammal.wide$ecosystem) <- c("G", "B", "C", "JS", "PJ")
```


How many bee and mammal species do we have?
```{r}
# Bee
n_distinct(bee.wide$species_code)

# Mammal
n_distinct(mammal.wide$species_code)
```



## Combine data

Need to ask: Do species_CV differ between bees and small mammals? 
```{r}
# Add a column to mammal data saying it's a mammal. Same for bee data
mammal.wide$species_type <- "mammal"
bee.wide$species_type <- "bee"

names(mammal.wide)
names(bee.wide)

# Merge with rbind
mammal.bee.data <- rbind(mammal.wide, bee.wide)

# Write csv for Kamren
write.csv(mammal.bee.data, "~/GitHub/SEV_Species_Interactions_Stability/data/mammal.bee.data.csv", row.names = F)

# Which ecosystems are in the final dataset for bees and for mammals?
mammal.bee.data %>%
  group_by(ecosystem, species_type) %>%
  summarise(n = n())
```




## Fit model

```{r}
### With ecosystem as a covariate (probably should have as a RE because only 3 levels). I checked, and including ecosystem as RE also gives a significant effect.
bm.mod <- lmer(species_CV ~ species_type + ecosystem + (1|species_code), data = mammal.bee.data %>% filter(ecosystem != "B"))
Anova(bm.mod)

# Check resids
plot(bm.mod)
plot(simulateResiduals(bm.mod))

# Plot the model prediction
bee.mammal.plot <- plot_model(bm.mod, type = "pred", terms = "species_type", color = "black") +
  scale_x_discrete(limits = c("bee", "mammal"),
                   labels = c("Bees", "Mammals")) +
  labs(y = "CV of Species Biomass", x = "Species Type")
bee.mammal.plot



### Logged for better residuals
bm.ln.mod <- lmer(log(species_CV) ~ species_type + ecosystem + (1|species_code), data = mammal.bee.data %>% filter(ecosystem != "B"))
Anova(bm.ln.mod)

# Check resids
plot(bm.ln.mod)
plot(simulateResiduals(bm.ln.mod))

# Plot the model prediction
bee.mammal.plot <- plot_model(bm.ln.mod, type = "pred", terms = "species_type", color = "black") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.05) +
  scale_x_discrete(limits = c("bee", "mammal"),
                   labels = c("Bees", "Rodents")) +
  labs(y = "Temporal Variability (Abundance CV)", x = "Species Type")
bee.mammal.plot

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/bee.mammal.plot.jpg", bee.mammal.plot, dpi = 600, width = 2100, height = 2500, units = "px")
```

### Try excluding blue grama site from bee data, and fit same models (just to make sure it doesn't change things). It doesn't!
```{r}
# What ecosystems are currently in the data?
unique(mammal.bee.data$ecosystem)

# Filter out blue grama site (ecosystem "B")
mammal.bee.data.noB <- mammal.bee.data %>%
  filter(ecosystem != "B")

# Make sure it's not there
unique(mammal.bee.data.noB$ecosystem)

# Fit model using new data
bm.noB.mod <- lmer(species_CV ~ species_type + ecosystem + (1|species_code), data = mammal.bee.data.noB)
Anova(bm.noB.mod)

# Also ln-transform species_CV
bm.noB.ln.mod <- lmer(log(species_CV) ~ species_type + ecosystem + (1|species_code), data = mammal.bee.data.noB)
Anova(bm.noB.ln.mod)

### Effect sizes (% increase in bee variability) - use raw data model for estimates

# Pollination
bm.noB.emm <- summary(emmeans(bm.noB.mod, pairwise ~ species_type))
(bm.noB.emm$emmeans[1, 2] - bm.noB.emm$emmeans[2, 2]) / bm.noB.emm$emmeans[2, 2] * 100
```

### Plot bee CV vs. plant CV in each ecosystem (just 3 ecosystems)

I need to give the sites a common name in mammal.bee.data.temp and the data.2012. 
```{r}
# Merge bee data with plant data
mammal.bee.data.temp <- mammal.bee.data.noB
colnames(mammal.bee.data.temp)[24] <- "bee_CV"

# Fix and finish...
catch <- left_join(data.2012, mammal.bee.data.temp)
```



# PICs - Pollination, Dispersal, Legume

## Prep for making a tree for this analysis

Prepping for tree prep in "Making_phylo_tree.Rmd"
```{r}
# Load species list data, so I can make "taxon.new" column from genus_species
sp.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/SevilletaSpeciesList2019.csv", stringsAsFactors = T)

sp.data$taxon.new <- paste(sp.data$genus, sp.data$sp_epithet, sep = "_")

sp.data %<>%
  dplyr::select(kartez, taxon.new)

data.2012.pic <- left_join(data.2012, sp.data, by = "kartez")
write.csv(data.2012.pic, "~/GitHub/SEV_Species_Interactions_Stability/data/data.2012.pic.csv", row.names = F) # For "Making_phylo_tree.Rmd"
```

## Prepping the data and the tree for PICs

```{r}
# Read in new PIC data with genera removed that aren't on Tim's megaphylogeny
PIC.data <- read.csv("~/GitHub/SEV_Species_Interactions_Stability/data/phylo_tree/new.pic.data.csv")

# Load tree
tr <- read.tree("~/GitHub/SEV_Species_Interactions_Stability/data/phylo_tree/Kamren.tre")

# Visualize tree 
par(mar = c(1, 1, 2, 1), + 0.1)
plot.phylo(tr, type = "phylogram", direction = "right", show.tip.label = T, use.edge.length = FALSE, cex = 0.6, label.offset = 1)
add.scale.bar(cex = 0.7)
nodelabels(tr$node.label, font = 2, bg = "white", frame = "r", cex = 0.5)

# This is the final tree Tim Farkas used
plotTree(tr)

# adjust branch lengths
tree2 <- compute.brlen(tr, method="Grafen")  # set branch lengths = Grafen
tree2$edge.length

plot(tree2)

# Check that tree and data have the same species -- Nope! 
# Note: setdiff returns items in the first set, but not in the second set.  Thus set order matters.
intersect(unique(PIC.data$taxon.new), tree2$tip.label)
setdiff(unique(PIC.data$taxon.new), tree2$tip.label)
setdiff(tree2$tip.label, unique(PIC.data$taxon.new)) 

#resolve multifurcations with branches of zero length
tree4 <- multi2di(tree2)   

# Save tree
write.nexus(tree4, file = "~/GitHub/SEV_Species_Interactions_Stability/data/phylo_tree/tree4.nex") #Nexus format

# Load libraries
libraries.jenn <- c("pez", "vegan", "picante", "lattice", "Hmisc", "readxl", "phytools", "visreg", "ggtree", "tidyverse", "U.Taxonstand")
invisible(lapply(libraries.jenn, require, character.only = T))
```

Prep data and tree (Tim’s tree “tr”) for PICs
```{r}
# adjust branch lengths
#tree2$edge.length<-rep(1,nrow(tree$edge))    # set branch lengths all = 1 
#tree2<-compute.brlen(tree2, 1)               # set branch lengths all = 1 alternate way
tree2 <- compute.brlen(tr, method="Grafen")  # set branch lengths = Grafen
tree2$edge.length

plot(tree2)

# Change binary variables to 0 and 1 (for pollinatoin & dispersal: 0 = wind, 1 = animal; for legume: 0 = not legume, 1 = legume)
PIC.data %<>%
  mutate(Pollination.new = ifelse(Pollination.new == "W", 0, 1),
         Dispersal.new = ifelse(Dispersal.new == "W", 0, 1),
         Legume = ifelse(Legume == "N", 0, 1)) 

# I need a dataset with taxon name as the rownames, and no repeat row names. So take the mean CV for each species so I have 1 row per species (if I use brms I won't need to do this)
data.pics.means <- PIC.data %>%
  group_by(taxon.new) %>%
  dplyr::summarise(cv.mean = mean(species_CV, na.rm=T),
            Pollination.new = mean(Pollination.new, na.rm=T),
            Dispersal.new = mean(Dispersal.new, na.rm=T),
            Legume = mean(Legume, na.rm=T)) %>%
  data.frame()

row.names(data.pics.means) <- data.pics.means$taxon.new # put taxa as row names (in same format as tree)

```


## Test for phylo signal in species interactions

```{r}
phylo.intrxn.names <- c("Pollination.new", "Dispersal.new", "Legume")  # May add back seed_mass"
phylo.intrxn <- match(phylo.intrxn.names, colnames(data.pics.means))

phylogsig.results <- list()

for (i in phylo.intrxn) {
  
  intrxn <- colnames(data.pics.means[i])
  data.pics.means$temp.data <- data.pics.means[, i]
  
  phylosig.lambda <- phylosig(tree4, data.pics.means$temp.data, method = "lambda", test = TRUE) # method can be either "lambda" or "K"
  phylosig.k <- phylosig(tree4, data.pics.means$temp.data, method = "K", test = TRUE) 
  
  phylogsig.results[[intrxn]][["lambda"]] <- phylosig.lambda
  phylogsig.results[[intrxn]][["K"]] <- phylosig.k
}

data.pics.means$temp.data <- NULL

phylogsig.results
```

## Run PICs

Calculate PICs and bind them together in single dataframe
```{r}
### BEFORE RUNNING THIS CODE: I need to remove species with missing data for each species interaction, and drop the tip from the tree. So make a separate tree for each species interaction, and don't do this in a loop.

is.na(data.pics.means$Pollination.new)
data.pics.means[75,]

### Pollination.new

# Remove missing data from df and tree
pic.pol.df <- data.pics.means %>% filter(!is.na(Pollination.new))
sum(is.na(pic.pol.df$Pollination.new))
tree.pol <- drop.tip(tree4, setdiff(tree4$tip.label, pic.pol.df$taxon.new))

# Calculate PIC
PIC.pol <- as.data.frame(pic(pic.pol.df$Pollination.new, tree.pol))
PIC.cv.pol <- as.data.frame(pic(pic.pol.df$cv.mean, tree.pol))
PIC.pol <- cbind(PIC.pol, PIC.cv.pol)
colnames(PIC.pol) <- c("Pollination.pic", "CV.pic")

# Fit PIC model
pic.pol.mod <- lm(CV.pic ~ 0 + Pollination.pic, data = PIC.pol, REML = F) # restricts line to go through origin
summary(pic.pol.mod)


### Dispersal.new

# Remove missing data from df and tree
pic.disp.df <- data.pics.means %>% filter(!is.na(Dispersal.new))
sum(is.na(pic.disp.df$Dispersal.new))
tree.disp <- drop.tip(tree4, setdiff(tree4$tip.label, pic.disp.df$taxon.new))

# Calculate PIC
PIC.disp <- as.data.frame(pic(pic.disp.df$Dispersal.new, tree.disp))
PIC.cv.disp <- as.data.frame(pic(pic.disp.df$cv.mean, tree.disp))
PIC.disp <- cbind(PIC.disp, PIC.cv.disp)
colnames(PIC.disp) <- c("Dispersal.pic", "CV.disp")

# Fit PIC model
pic.disp.mod <- lm(CV.disp ~ 0 + Dispersal.pic, data = PIC.disp, REML = F) # restricts line to go through origin
summary(pic.disp.mod)



### Legume

# Remove missing data from df and tree
pic.leg.df <- data.pics.means %>% filter(!is.na(Legume))
sum(is.na(pic.leg.df$Legume))
tree.leg <- drop.tip(tree4, setdiff(tree4$tip.label, pic.leg.df$taxon.new))

# Calculate PIC
PIC.leg <- as.data.frame(pic(pic.leg.df$Legume, tree.leg))
PIC.cv.leg <- as.data.frame(pic(pic.leg.df$cv.mean, tree.leg))
PIC.leg <- cbind(PIC.leg, PIC.cv.leg)
colnames(PIC.leg) <- c("Legume.pic", "CV.leg")

# Fit PIC model
pic.leg.mod <- lm(CV.leg ~ 0 + Legume.pic, data = PIC.leg, REML = F) # restricts line to go through origin
summary(pic.leg.mod)
```



# brms models

## Pollination

```{r}
# Import new.pic.data (created in "Making_phylo_tree.Rmd) 
new.pic.data <- read.csv("./data/phylo_tree/new.pic.data.csv")

# Construct a covariance matrix of species
A.init <- ape::vcv.phylo(tree4)

intersect(new.pic.data$taxon.new, rownames(A.init)) # 151 are good
setdiff(new.pic.data$taxon.new, rownames(A.init))
setdiff(rownames(A.init), new.pic.data$taxon.new)

# Save and read in phylogenetic data
phylo_list_init <- list(A.init = A.init)
dput(phylo_list_init, "phylo_list_init.txt")
phylo_list_init_r <-dget("phylo_list_init.txt")
```

Fit model. Suppressed for run time. 
```{r, eval=F}
library(brms)
library(tidybayes)

# With quad_ID. REPLACE WITH LAT/LONG WHEN I GET CODE FROM ARA.
#brms.mod.pol <- bf(species_CV ~ (1|Pollination.new) + (1|quad_ID) + (1|site) + (1|gr(taxon.new, cov = A.init)))
brms.mod.pol <- bf(species_CV ~ (1|Pollination.new) + (1|gr(taxon.new, cov = A.init)))
brms.mod.pol

# Set our own priors for pollination model
prior.pol <- c(
            set_prior("student_t(3, 0, 1)", class = "sd", group = "Pollination.new") #, # Try normal(0, 1) instead of student_t
            #set_prior("student_t(3, 0, 1)", class = "sd", group = "quad_ID"),
            #set_prior("student_t(3, 0, 1)", class = "sd", group = "site")
            )
prior.pol

# Run the model
k_fit_brms.pol <- brm(brms.mod.pol, 
                  data = new.pic.data, data2 = phylo_list_init_r,
                  family = "gaussian", prior = prior.pol,
                  chains = 4, cores = 8, iter = 2000, backend = "cmdstanr",
                  control = list(adapt_delta = 0.85, max_treedepth = 12)) # step_size = 0.01

summary(k_fit_brms.pol, prob = 0.95) # 45 divergent transitions

#Multilevel Hyperparameters:
#~Pollination.new (Number of levels: 2) 
#              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.43      0.49     0.01     1.83 1.00     1762     1956
#
#~taxon.new (Number of levels: 126) 
#              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     2.87      0.23     2.46     3.35 1.01      772     1561

#Regression Coefficients:
#          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept     2.27      1.21    -0.07     4.63 1.00     1015     1795

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     0.74      0.01     0.73     0.75 1.00     7753     2743

summary(k_fit_brms.pol, prob = 0.5) # 45 divergent transitions

#Multilevel Hyperparameters:
#~Pollination.new (Number of levels: 2) 
#              Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.43      0.49     0.11     0.61 1.00     1762     1956

#~taxon.new (Number of levels: 126) 
#              Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     2.87      0.23     2.72     3.02 1.01      772     1561

#Regression Coefficients:
#          Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#Intercept     2.27      1.21     1.48     3.05 1.00     1015     1795

#Further Distributional Parameters:
#      Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sigma     0.74      0.01     0.73     0.74 1.00     7753     2743


# This code pulls out the sd and squares it to give the variance explained
# from https://discourse.mc-stan.org/t/calculating-variance-explained-by-each-predictor-and-clustering-variable-in-brms/11223/7
# written by Jeffrey Girard
k_fit_brms.pol %>% 
  tidy_draws() %>% 
  select(starts_with("sd_"), sigma) %>% 
  transmute_all(.funs = list(sq = ~(. ^ 2))) %>% 
  mutate(total_var = rowSums(.)) %>% 
  mutate_at(.vars = vars(-total_var), 
            .funs = list(pct = ~(. / total_var))) %>% 
  map_df(.f = ~ median_hdci(., .width = 0.5), .id = "component") %>% 
  mutate(
    component = str_remove(component, "sd_"),
    component = str_remove(component, "__Intercept_sq"),
    component = str_replace(component, "sigma_sq", "residual")
  )

#             component            y         ymin         ymax .width .point .interval
#1     Pollination.new 0.069789492 2.184011e-08 0.069789494    0.5 median      hdci
#2           taxon.new 8.176768852 7.057258e+00 8.772081576    0.5 median      hdci
#3            residual 0.546527526 5.392215e-01 0.552158969    0.5 median      hdci
#4           total_var 9.049949979 7.906783e+00 9.758757537    0.5 median      hdci
#5 Pollination.new_pct 0.007761887 2.492564e-09 0.007761889    0.5 median      hdci
#6       taxon.new_pct 0.925289632 9.190998e-01 0.943664421    0.5 median      hdci
#7        residual_pct 0.060307420 5.289650e-02 0.065787905    0.5 median      hdci

plot(k_fit_brms.pol)

### Model checking plots. Posterior predict. 
pp_check(k_fit_brms.pol, resp = "species_CV", ndraws = 100)

# Full group level estimates for random effects

ranef(k_fit_brms.pol, groups = "Pollination.new", probs = 0.95) 

#     Estimate Est.Error       Q95
#A -0.03392375 0.3879865 0.5420103
#W  0.02167897 0.3855266 0.6227592

ranef(k_fit_brms.pol, groups = "site", probs = 0.95)
```

Example code to get numbers to report for categorical predictor in brms:
```{r, eval=F}
### To get level-specific estimates
nd <- data.frame(x = c("A","B"))
ep <- posterior_epred(k_fit_brms.pol, newdata = nd, re_formula = NA)   # draws on outcome scale
# columns: [,1] = level A, [,2] = level B

summ <- function(z) c(
  mean = mean(z),
  l95 = quantile(z, 0.025),
  u95 = quantile(z, 0.975)
)

level_A <- summ(ep[,1])
level_B <- summ(ep[,2])

### To report the contrast:
diff_BA <- ep[,2] - ep[,1]
diff_summary <- summ(diff_BA)

# Optional: posterior probability of direction (how sure we are the difference is > 0)
pd <- max(mean(diff_BA > 0), mean(diff_BA < 0))  # e.g., 0.97 ~ "97% probability the effect is positive"
```




Try model without phylogeny. Does it detect a difference then? 
```{r,eval=F}
# With quad_ID. REPLACE WITH LAT/LONG WHEN I GET CODE FROM ARA.
brms.mod.pol2 <- bf(species_CV ~ (1|Pollination.new) + (1|quad_ID))
brms.mod.pol2

# Set our own priors for pollination model
prior.pol2 <- c(
            set_prior("student_t(3, 0, 1)", class = "sd", group = "Pollination.new"),
            set_prior("student_t(3, 0, 1)", class = "sd", group = "quad_ID"))
prior.pol2

# Run the model
k_fit_brms.pol2 <- brm(brms.mod.pol2, 
                  data = new.pic.data, data2 = phylo_list_init_r,
                  family = "gaussian", prior = prior.pol2,
                  chains = 4, cores = 8, iter = 2000, backend = "cmdstanr",
                  control = list(adapt_delta = 0.85, max_treedepth = 12))

summary(k_fit_brms.pol2, prob = 0.95) 
#Multilevel Hyperparameters:
#~Pollination.new (Number of levels: 2) 
#              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.69      0.42     0.19     1.80 1.00     2099     2163
#
#~quad_ID (Number of levels: 451) 
#              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.23      0.01     0.21     0.26 1.00     1587     2506
#
#Regression Coefficients:
#          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept     2.22      0.50     1.16     3.34 1.01     1781     1564

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     0.87      0.01     0.86     0.88 1.00     6901     2589

summary(k_fit_brms.pol2, prob = 0.5) 

#Multilevel Hyperparameters:
#~Pollination.new (Number of levels: 2) 
#              Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.69      0.42     0.38     0.90 1.00     2099     2163
#
#~quad_ID (Number of levels: 451) 
#              Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sd(Intercept)     0.23      0.01     0.22     0.24 1.00     1587     2506
#
#Regression Coefficients:
#          Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#Intercept     2.22      0.50     1.96     2.46 1.01     1781     1564
#
#Further Distributional Parameters:
#      Estimate Est.Error l-50% CI u-50% CI Rhat Bulk_ESS Tail_ESS
#sigma     0.87      0.01     0.86     0.87 1.00     6901     2589


# This code pulls out the sd and squares it to give the variance explained
# from https://discourse.mc-stan.org/t/calculating-variance-explained-by-each-predictor-and-clustering-variable-in-brms/11223/7
# written by Jeffrey Girard
k_fit_brms.pol2 %>% 
  tidy_draws() %>% 
  select(starts_with("sd_"), sigma) %>% 
  transmute_all(.funs = list(sq = ~(. ^ 2))) %>% 
  mutate(total_var = rowSums(.)) %>% 
  mutate_at(.vars = vars(-total_var), 
            .funs = list(pct = ~(. / total_var))) %>% 
  map_df(.f = ~ median_hdci(., .width = 0.5), .id = "component") %>% 
  mutate(
    component = str_remove(component, "sd_"),
    component = str_remove(component, "__Intercept_sq"),
    component = str_replace(component, "sigma_sq", "residual")
  )

#            component          y       ymin       ymax .width .point .interval
#1     Pollination.new 0.33584005 0.02138701 0.33934703    0.5 median      hdci
#2             quad_ID 0.05306965 0.04810389 0.05614957    0.5 median      hdci
#3            residual 0.75318879 0.74563743 0.75919680    0.5 median      hdci
#4           total_var 1.14407411 0.82536581 1.15116110    0.5 median      hdci
#5 Pollination.new_pct 0.29360702 0.03925617 0.30723646    0.5 median      hdci
#6         quad_ID_pct 0.04566902 0.03915803 0.06051703    0.5 median      hdci
#7        residual_pct 0.65872852 0.64248456 0.89370851    0.5 median      hdci

ranef(k_fit_brms.pol2, groups = "Pollination.new", probs = 0.95) 

#    Estimate Est.Error       Q95
#A  0.2714804 0.4994677 1.0904480
#W -0.2699768 0.4993902 0.5409442
```

## Dispersal

```{r, eval=F}
# With quad_ID. REPLACE WITH LAT/LONG WHEN I GET CODE FROM ARA.
#brms.mod.pol <- bf(species_CV ~ (1|Pollination.new) + (1|quad_ID) + (1|site) + (1|gr(taxon.new, cov = A.init)))
brms.mod.disp <- bf(species_CV ~ (1|Dispersal.new) + (1|gr(taxon.new, cov = A.init)))
brms.mod.disp

# Set our own priors for dispersal model
prior.disp <- c(
            set_prior("student_t(3, 0, 1)", class = "sd", group = "Dispersal.new") #, # Try normal(0, 1) instead of student_t
            #set_prior("student_t(3, 0, 1)", class = "sd", group = "quad_ID"),
            #set_prior("student_t(3, 0, 1)", class = "sd", group = "site")
            )
prior.disp

# Run the model
k_fit_brms.disp <- brm(brms.mod.disp, 
                  data = new.pic.data, data2 = phylo_list_init_r,
                  family = "gaussian", prior = prior.disp,
                  chains = 4, cores = 8, iter = 2000, backend = "cmdstanr",
                  control = list(adapt_delta = 0.85, max_treedepth = 12)) # step_size = 0.01

summary(k_fit_brms.disp, prob = 0.95) 

summary(k_fit_brms.disp, prob = 0.5) 

# This code pulls out the sd and squares it to give the variance explained
# from https://discourse.mc-stan.org/t/calculating-variance-explained-by-each-predictor-and-clustering-variable-in-brms/11223/7
# written by Jeffrey Girard
k_fit_brms.disp %>% 
  tidy_draws() %>% 
  select(starts_with("sd_"), sigma) %>% 
  transmute_all(.funs = list(sq = ~(. ^ 2))) %>% 
  mutate(total_var = rowSums(.)) %>% 
  mutate_at(.vars = vars(-total_var), 
            .funs = list(pct = ~(. / total_var))) %>% 
  map_df(.f = ~ median_hdci(., .width = 0.5), .id = "component") %>% 
  mutate(
    component = str_remove(component, "sd_"),
    component = str_remove(component, "__Intercept_sq"),
    component = str_replace(component, "sigma_sq", "residual")
  )

plot(k_fit_brms.disp)

### Model checking plots. Posterior predict. 
pp_check(k_fit_brms.disp, resp = "species_CV", ndraws = 100)

# Full group level estimates for random effects
ranef(k_fit_brms.disp, groups = "Dispersal.new", probs = 0.95) 

```






# Make a phylogenetic tree with traits on it

```{r, message=F}
library(phylobase)
library(adephylo)
library(RColorBrewer)
```


```{r}
length(tree4$tip.label)
length(unique(tree4$tip.label))

### Create a df of plants (labeled with their taxon name) and their CV and Pollination Strategy

# First get mean of CV for each species. I think this is the best way to get a single CV value for each species (averaged across quads)
new.pic.data.summary <- new.pic.data %>%
  group_by(kartez) %>%
  summarise(n = n(),
            mean.cv = mean(species_CV, na.rm=T))

phylo.data <- left_join(new.pic.data.summary, new.pic.data %>% dplyr::select(kartez, Pollination.new, Dispersal.new, taxon.new), by = "kartez") %>% group_by(kartez) %>% filter(row_number()==1) %>% data.frame()

# Change "taxon.new" to "label" so I can merge with tree data below
colnames(phylo.data)[6] <- "label"

# Row names as genus species
rownames(phylo.data) <- phylo.data$label

# Remove columns I don't need
phylo.data$kartez <- NULL
#phylo.data$taxon.new <- NULL
phylo.data$n <- NULL

colnames(phylo.data)
```

Trying with ggtree
```{r, message=F}
library(treeio)      # for importing trees
library(ggnewscale)  # to use multiple color scales
```


```{r}
# Join data to tree
p <- ggtree(tree4, layout="fan") # This makes the tree circular

tree4$tip.label.genus <- sub("_.*", "", tree4$tip.label)
#tree4$tip.label.genus <- append(tree4$tip.label.genus, rep(NA, 150))
tree4$tip.label.genus <- append(tree4$tip.label.genus, rep(NA, 126))

p$data$tip.label.genus <- tree4$tip.label.genus

# Playing with having only one label per genus
# Make sure tips have genus info
# (here assuming tip.label.genus exists in d)
# If not, you can join it in from a metadata table
#d <- p$data
#stopifnot("tip.label.genus" %in% colnames(d))

# Loop over each genus and add a clade label
#for (g in unique(d$tip.label.genus[!is.na(d$tip.label.genus)])) {
#  genus_tips <- d$label[d$tip.label.genus == g & d$isTip]
#  mrca_node <- MRCA(tree4, genus_tips)
#  p <- p + geom_cladelabel(node = mrca_node, 
#                           label = g, 
#                           align = TRUE, 
#                           barsize = 1,
#                           fontsize = 3)
#}



# Extract tip positions for label/dot placement
tip_positions <- p$data %>%
  filter(isTip) %>%
  select(label, x, y)

# Merge with  metadata
tip_data_merged <- tip_positions %>%
  left_join(phylo.data, by = "label")

p <- p + geom_tiplab(aes(label = tip.label.genus), size = 2) + # tip labels on tips
  xlim(0, max(tip_positions$x) + 0.7)   # extend plot space for dots

# Base tree with tip labels
#p1 <- p +
#  geom_tiplab(size = 2, hjust = 0) +  # tip labels on tips
#  xlim(0, max(tip_positions$x) + 0.5)   # extend plot space for dots

# Add first color scale: continuous variable
p1 <- p +
  geom_point(data = tip_data_merged,
             aes(x = x + 0.7, y = y, color = mean.cv),
             size = 2) +
  scale_color_viridis_c(name = "Biomass CV") +
  ggnewscale::new_scale_color()

### Add second color scale: categorical variable 

# Pollination
p.pol <- p1 +
  geom_point(data = tip_data_merged,
             aes(x = x + 0.6, y = y, color = Pollination.new),
             size = 2) +
  scale_color_manual(values = c("W" = "blue", "A" = "red"),
                     name = "Pollination Strategy",
                     labels = c("Animal", "Wind")) +
  theme_tree2()

p.pol

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/tree_with_pollination.png", p.pol, dpi = 600, width = 4000, height = 6000, units = "px")

# Dispersal
p.disp <- p1 +
  geom_point(data = tip_data_merged,
             aes(x = x + 0.6, y = y, color = Dispersal.new),
             size = 2) +
  scale_color_manual(values = c("W" = "darkgreen", "A" = "orange"),
                     name = "Dispersal Strategy",
                     labels = c("Animal", "Wind")) +
  theme_tree2()

p.disp

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/tree_with_dispersal.png", p.disp, dpi = 600, width = 4000, height = 6000, units = "px")
```




Old attempts
```{r, eval=F}
### 1

# Join data to tree
p <- ggtree(tree4) %<+% tip_data

# Plot with continuous variable as tip color
p1 <- p + 
  geom_tippoint(aes(color = species_CV), size = 3) +
  scale_color_viridis_c(name = "Species CV") +
  ggnewscale::new_scale_color() +

  # Categorical variable point (e.g. right point)
  geom_tippoint(aes(x = x + 0.5, color = pollination.new), size = 3) +
  scale_color_manual(values = c("wind" = "blue", "insect" = "green", "bird" = "red"),
                     name = "Pollination Type") +
  
  # Add tip labels to the right of both dots
  geom_tiplab(aes(x = x + 1.5), hjust = 0) +
  theme_tree2()

p1


### 2 

# Visualize the tree with traits
mypalette <- colorRampPalette(brewer.pal(9, "RdBu"))

par(mar = c(1, 1, 1, 1) + 0.1)
x <- phylo4d(tree4, phylo.data)

png(filename = "~/GitHub/SEV_Species_Interactions_Stability/reports/tree_with_traits.png",  width = 55, 
    height = 70, units = "cm", res = 300)

table.phylo4d(x, treetype = "phylo", symbol = "colors", show.node = TRUE, cex.label = 1, scale = F, use.edge.length = F, edge.color = "black", edge.width = 2, box = F, col = mypalette(25), pch = 15, cex.symbol = 1.25, ratio.tree = .9, cex.legend = 1.5, center = F)

dev.off()

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/tree_with_traits.png", tree_with_traits, dpi = 600, width = 2000, height = 2500, units = "px")

```













# Code graveyard

This code took data from the website and processed it to add 0's and NAs where appropriate. However, my computer couldn't run this code. So Jenn ran it (and touched it up -- see her email from 4/1/2025). Now I simply import the csv file she sent me.
```{r, eval=F, echo=F}
options(HTTPUserAgent="EDI_CodeGen")
          
inUrl1  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-sev/331/3/cb45bdfc2edb50ef242d29f16055ab5b" 
infile1 <- tempfile()
try(download.file(inUrl1,infile1,method="curl",extra=paste0(' -A "',getOption("HTTPUserAgent"),'"')))
if (is.na(file.size(infile1))) download.file(inUrl1,infile1,method="auto")

# This is the data with CV info
mass <-read.csv(infile1,header=F 
                , stringsAsFactors = T
                ,skip=1
                ,sep=","  
                ,quot='"' 
                , col.names=c(
                    "site",     
                    "year",     
                    "season",     
                    "date",     
                    "web",     
                    "transect",     
                    "block",     
                    "plot",     
                    "subplot",     
                    "quad",     
                    "treatment",     
                    "kartez",     
                    "genus",     
                    "sp.epithet",     
                    "family",     
                    "LifeHistory",     
                    "PhotoPath",     
                    "FunctionalGroup",     
                    "cover",     
                    "volume",     
                    "biomass.BM",     
                    "biomass.BIM",     
                    "SiteCluster",     
                    "MetStation",     
                    "season.precip",     
                    "GDD",     
                    "SPEI.comp"    ), check.names=TRUE)
               
unlink(infile1)


mass$web <- as.factor(mass$web)

### recode treatment into three columns for factorial 2 x 2 x 2 design (NOTE: BELOW I'LL NEED TO SUBSET TO treatment == C, N, P, T, TN, PN, TP, TPN)
# note 'T' requires quotes so it is not read as TRUE
mass$warm <- recode_factor(mass$treatment, C="Ambient_T",N="Ambient_T",P="Ambient_T",'T'="Warmed",TN="Warmed",PN="Ambient_T",TP="Warmed",TPN="Warmed")

mass$ppt <- recode_factor(mass$treatment, C="Ambient_ppt",N="Ambient_ppt",P="Extra_winter_ppt",'T'="Ambient_ppt",TN="Ambient_ppt",PN="Extra_winter_ppt",TP="Extra_winter_ppt",TPN="Extra_winter_ppt")

mass$nitrogen <- recode_factor(mass$treatment, C="Ambient_N",N="N_addition",P="Ambient_N",'T'="Ambient_N",TN="N_addition",PN="N_addition",TP="Ambient_N",TPN="N_addition")

# Make unique quad ID factor
mass$quad_ID <- as.factor(
      paste(
        mass$site,
        mass$web,
        mass$transect,
        mass$treatment,
        mass$block,
        mass$plot,
        mass$subplot,
        mass$quad,
        sep = "_"))

# cast the data to create columns for each plant species so zeros are added for dates where species was absent  --------------------- for all quads
mass_species <- dcast(mass, year + site + season + season.precip + GDD + SPEI.comp + SiteCluster + MetStation + web + transect + treatment + block + plot + subplot + quad + quad_ID ~ kartez, sum, value.var="biomass.BM", fill=0)

#get rid of column called EMPTY - this was a placeholder for quadrats that had bare ground, not a real species
mass_species <- mass_species %>% select(-EMPTY)

# calculate total biomass
mass_species$totmass <- rowSums(mass_species[,17:310])

# reshape again so each species in each quadrat is a row, cols are timepoints
# create a continuous time variable
mass_species$seas <- as.numeric(as.character(recode_factor(mass_species$season, fall="0.5", spring="0")))
mass_species$year_seas <- as.factor(mass_species$year+mass_species$seas)                           
mass_species$time <- as.numeric(as.factor(mass_species$year_seas))
mass_species$year.f <- as.factor(mass_species$year)

###### total biomass stability ----------------------
#tot mass
totmass <- mass_species %>% select(year.f,site,season,seas,year_seas,season.precip,GDD,SPEI.comp, SiteCluster,MetStation,web,transect,treatment,block,plot,subplot,quad,quad_ID,totmass)

totmass_cast <- dcast(totmass,site+season+seas+SiteCluster+MetStation+web+transect+treatment+block+plot+subplot+quad+quad_ID ~ year.f , sum, value.var="totmass",fill=-9999)
# fix problem with fill in dcast
totmass_cast[totmass_cast== "-9999"] <- "NA"
summary(totmass_cast)
totmass_cast <- totmass_cast %>% mutate_if(is.character, as.numeric)

# calculate totmass sd
totmass_cast$sd<-transform(totmass_cast, totmass_stdev=apply(totmass_cast[,14:38], 1, sd, na.rm=TRUE))

# Melt data to make a row for each species in each quad. To do this properly, I need every column name that isn't a species listed in the "id" argument.
mass_species$block <- as.factor(mass_species$block)
mass_species_melt <- melt(mass_species, id = c("year", "site", "season", "season.precip", "GDD", "SPEI.comp", "SiteCluster", "MetStation", "web", "transect", "treatment", 
"block", "plot", "subplot", "quad", "quad_ID", "totmass", "seas", "year_seas", 
"time", "year.f"), variable.name = "kartez", value.name = "biomass.BM")

mass_species_melt$biomass.BM <- as.numeric(mass_species_melt$biomass.BM)
summary(mass_species_melt$biomass.BM)
summary(mass_species_melt$kartez)

# Cast to make a column for each year_seas, and fill with the sum of biomass (rows are species x quad). This is freezing up R, not sure what to do about it...

#mass_species_cast <- dcast(mass_species_melt, year + site + season + season.precip + GDD + SPEI.comp + SiteCluster + MetStation + web + transect + treatment + block + plot + subplot + quad + quad_ID + totmass + seas + time + year.f + kartez ~ year_seas, sum, value.var = "biomass.BM", fill="NA")

#write.csv(mass_species_cast, "mass_species_cast.csv")
```




## 2004 data

### 2004: Model excluding WA (for both pollination and dispersal)

```{r, eval=F}
# I need to change the order of "Legume" levels for the plot
data.2004$Legume <- factor(data.2004$Legume, levels = c("Y", "N"))

### Full model ###
mod.all.2004.f3.gps <- lmer(species_CV ~ Pollination + Dispersal + Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Pollination != "WA" & Dispersal != "WA"))
Anova(mod.all.2004.f3.gps, type = 2)

### Separate model for each predictor ###
mod.pol.2004.f3.gps <- lmer(species_CV ~ Pollination + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Pollination != "WA"))
mod.disp.2004.f3.gps <- lmer(species_CV ~ Dispersal + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Dispersal != "WA"))
mod.leg.2004.f3.gps <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004)

Anova(mod.pol.2004.f3.gps, type = 2)
Anova(mod.disp.2004.f3.gps, type = 2)
Anova(mod.leg.2004.f3.gps, type = 2)

# Plot marginal effect of each predictor
plot_model(mod.pol.2004.f3.gps, type = "pred", terms = "Pollination")
plot_model(mod.disp.2004.f3.gps, type = "pred", terms = "Dispersal")
plot_model(mod.leg.2004.f3.gps, type = "pred", terms = "Legume")

### Fit model with Pollination x Dispersal
mod.poldisp.2004.f3.gps <- lmer(species_CV ~ Pollination * Dispersal + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Pollination != "WA" & Dispersal != "WA"))
Anova(mod.poldisp.2004.f3.gps, type = 2)
```

Try models that include species_mu
```{r, eval=F}
### Separate model for each predictor ###
mod.pol.2004.f3.gps.mu <- lmer(species_CV ~ Pollination * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Pollination != "WA"))
mod.disp.2004.f3.gps.mu <- lmer(species_CV ~ Dispersal * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Dispersal != "WA"))
mod.leg.2004.f3.gps.mu <- lmer(species_CV ~ Legume * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2004)

Anova(mod.pol.2004.f3.gps.mu, type = 2)
Anova(mod.disp.2004.f3.gps.mu, type = 2)
Anova(mod.leg.2004.f3.gps.mu, type = 2)

# Plot marginal effect of each predictor
plot_model(mod.pol.2004.f3.gps.mu, type = "pred", terms = "Pollination")
plot_model(mod.disp.2004.f3.gps.mu, type = "pred", terms = "Dispersal")
plot_model(mod.leg.2004.f3.gps.mu, type = "pred", terms = "Legume")
```

### 2004: Simplified categories for pollination and dispersal

Here, species that are animal and wind pollinated are treated as wind pollinated, because they don’t rely on another species if they can be pollinated by wind.
```{r, eval=F}
### Full model ###
mod.all.2004.f2.gps <- lmer(species_CV ~ Pollination.new + Dispersal.new + Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004)
Anova(mod.all.2004.f2.gps, type = 2)

### Separate model for each predictor
mod.pol.2004.f2.gps <- lmer(species_CV ~ Pollination.new + X + Y + (1|kartez), na.action = na.omit, data = data.2004)
mod.disp.2004.f2.gps <- lmer(species_CV ~ Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = data.2004)
mod.leg.2004.f2.gps <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004)

Anova(mod.pol.2004.f2.gps, type = 2)
Anova(mod.disp.2004.f2.gps, type = 2)
Anova(mod.leg.2004.f2.gps, type = 2)

# Plot marginal effect of each predictor
plot_model(mod.pol.2004.f2.gps, type = "pred", terms = "Pollination.new")
plot_model(mod.disp.2004.f2.gps, type = "pred", terms = "Dispersal.new")
plot_model(mod.leg.2004.f2.gps, type = "pred", terms = "Legume")
```

### Make dfs for other years as cutoff

```{r, eval=F}
data.2006 <- data %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ", "fertilizer", "mixed_grass", "mixed_shrub", "MRME", "warming")) %>%
  dplyr::select(-c("X1999", "X2000", "X2001", "X2002", "X2003", "X2004", "X2005")) # Remove years that aren' included in this dataset, so I can calculate species_CV and species_mu across these years (everybody should have same time series)

data.2008 <- data %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ", "fertilizer", "mixed_grass", "mixed_shrub", "MRME", "warming", "NutNet")) %>%
  dplyr::select(-c("X1999", "X2000", "X2001", "X2002", "X2003", "X2004", "X2005", "X2006", "X2007")) # Remove years that aren' included in this dataset, so I can calculate species_CV and species_mu across these years (everybody should have same time series)

data.2011 <- data %>%
  filter(site %in% c("core_black", "core_blue", "core_creosote", "core_PJ", "fertilizer", "mixed_grass", "mixed_shrub", "MRME", "warming", "NutNet", "tower_east", "tower_west")) %>%
  dplyr::select(-c("X1999", "X2000", "X2001", "X2002", "X2003", "X2004", "X2005", "X2006", "X2007", "X2008", "X2009", "X2010")) # Remove years that aren' included in this dataset, so I can calculate species_CV and species_mu across these years (everybody should have same time series)
```

## Looping through 2004 and 2012 (not sure if I need this)

I think 2012 is the best option, but I'll loop through both datasets
```{r, eval=F}
final.df.list <- list(data.2004, data.2012)
final.df.names <- list("data.2004", "data.2012")
year.df <- list("2004", "2012")

model.outputs <- list()

for (i in 1:length(final.df.list)) {
  
  temp.df <- final.df.list[[i]]
  temp.df.name <- final.df.names[[i]]
  year <- year.df[[i]]
  temp.df$Legume <- factor(temp.df$Legume, levels = c("Y", "N"))  # Change the order of "Legume" levels for the plot
  
  ### Fit model using original predictors (including "WA" for pollination and dispersal)
  
  # Full model
  mod.all.orig <- lmer(species_CV ~ Pollination + Dispersal + Legume + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  # Individual models for each predictor
  mod.pol.orig <- lmer(species_CV ~ Pollination + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  mod.disp.orig <- lmer(species_CV ~ Dispersal + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  mod.leg.orig <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  # Model Pollination x Dispersal interaction
  mod.poldisp.orig <- lmer(species_CV ~ Pollination * Dispersal + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  model.outputs[[year]][["Original"]][["Full"]] <- Anova(mod.all.orig, type = 2)
  model.outputs[[year]][["Original"]][["Pollination"]] <- Anova(mod.pol.orig, type = 2)
  model.outputs[[year]][["Original"]][["Dispersal"]] <- Anova(mod.disp.orig, type = 2)
  model.outputs[[year]][["Original"]][["Legume"]] <- Anova(mod.leg.orig, type = 2)
  model.outputs[[year]][["Original"]][["PollinationxDispersal"]] <- Anova(mod.poldisp.orig, type = 2)
  
  ### Fit model excluding WA (for both pollination and dispersal)
  
  # Full model
  mod.all.noWA <- lmer(species_CV ~ Pollination + Dispersal + Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004 %>% filter(Pollination != "WA" & Dispersal != "WA"))
  
  # Individual models for each predictor
  mod.pol.noWA <- lmer(species_CV ~ Pollination + X + Y + (1|kartez), na.action = na.omit, data = temp.df %>% filter(Pollination != "WA"))
  mod.disp.noWA <- lmer(species_CV ~ Dispersal + X + Y + (1|kartez), na.action = na.omit, data = temp.df %>% filter(Dispersal != "WA"))
  mod.leg.noWA <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  # Model Pollination x Dispersal interaction
  mod.poldisp.noWA <- lmer(species_CV ~ Pollination * Dispersal + X + Y + (1|kartez), na.action = na.omit, data = temp.df %>% filter(Pollination != "WA" & Dispersal != "WA"))
  
  model.outputs[[year]][["NoWA"]][["Full"]] <- Anova(mod.all.noWA, type = 2)
  model.outputs[[year]][["NoWA"]][["Pollination"]] <- Anova(mod.pol.noWA, type = 2)
  model.outputs[[year]][["NoWA"]][["Dispersal"]] <- Anova(mod.disp.noWA, type = 2)
  model.outputs[[year]][["NoWA"]][["Legume"]] <- Anova(mod.leg.noWA, type = 2)
  model.outputs[[year]][["NoWA"]][["PollinationxDispersal"]] <- Anova(mod.poldisp.noWA, type = 2)
  
  ### Fit simplified categories for pollination and dispersal (WA is recast as W)
  
  # Full model
  mod.all.simp <- lmer(species_CV ~ Pollination.new + Dispersal.new + Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2004)

  # Separate model for each predictor
  mod.pol.simp <- lmer(species_CV ~ Pollination.new + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  mod.disp.simp <- lmer(species_CV ~ Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  mod.leg.simp <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  # Model Pollination x Dispersal interaction
  mod.poldisp.simp <- lmer(species_CV ~ Pollination.new * Dispersal.new + X + Y + (1|kartez), na.action = na.omit, data = temp.df)
  
  model.outputs[[year]][["Simplified"]][["Full"]] <- Anova(mod.all.simp, type = 2)
  model.outputs[[year]][["Simplified"]][["Pollination"]] <- Anova(mod.pol.simp, type = 2)
  model.outputs[[year]][["Simplified"]][["Dispersal"]] <- Anova(mod.disp.simp, type = 2)
  model.outputs[[year]][["Simplified"]][["Legume"]] <- Anova(mod.leg.simp, type = 2)
  model.outputs[[year]][["Simplified"]][["PollinationxDispersal"]] <- Anova(mod.poldisp.simp, type = 2)

}

model.outputs

### Plot marginal effect of each predictor - I need to save the models in the loop, and plot them in the loop perhaps? Or just plot them here as needed? 

#plot_model(mod.pol.noWA, type = "pred", terms = "Pollination")
#plot_model(mod.disp.noWA, type = "pred", terms = "Dispersal")
#plot_model(mod.leg.noWA, type = "pred", terms = "Legume")

#plot_model(mod.pol.simp, type = "pred", terms = "Pollination.new")
#plot_model(mod.disp.simp, type = "pred", terms = "Dispersal.new")
#plot_model(mod.leg.simp, type = "pred", terms = "Legume.new")
```


### 2012: Model excluding WA (for both pollination and dispersal)

```{r, eval=F}
# I need to change the order of "Legume" levels for the plot
data.2012$Legume <- factor(data.2012$Legume, levels = c("Y", "N"))

### Full model ###
mod.all <- lmer(species_CV ~ Pollination + Dispersal + Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA" & Dispersal != "WA"))
Anova(mod.all, type = 2)
plot(mod.all)

### Separate model for each predictor ###
mod.pol <- lmer(species_CV ~ Pollination + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA"))
mod.disp <- lmer(species_CV ~ Dispersal + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Dispersal != "WA"))
mod.leg <- lmer(species_CV ~ Legume + X + Y + (1|kartez), na.action = na.omit, data = data.2012)

Anova(mod.pol, type = 2)
Anova(mod.disp, type = 2)
Anova(mod.leg, type = 2)

# Plot marginal effect of each predictor
plot_model(mod.pol, type = "pred", terms = "Pollination")
plot_model(mod.disp, type = "pred", terms = "Dispersal")
plot_model(mod.leg, type = "pred", terms = "Legume")


# Plot marginal effect of each predictor
Pol.plot <- plot_model(mod.pol, type = "pred", terms = "Pollination") +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Pollination Strategy") +
  ylim(c(2.2, 3))

Pol.plot


Disp.plot <- plot_model(mod.disp, type = "pred", terms = "Dispersal")  +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Dispersal Strategy") +
  ylim(c(2.2, 3))

Disp.plot


Leg.plot <- plot_model(mod.leg, type = "pred", terms = "Legume")  +
  scale_x_discrete(limits = c("Y", "N"),
                   labels = c("Legume", "Not Legume")) +
  labs(y = "CV of Species Biomass", x = "Legume Status") +
  ylim(c(2.2, 3.1))

Leg.plot


ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Pol.plot.jpg", Pol.plot, dpi = 600, width = 2000, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Disp.plot.jpg", Disp.plot, dpi = 600, width = 2000, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Leg.plot.jpg", Leg.plot, dpi = 600, width = 2000, height = 2500, units = "px")







### Fit model with Pollination x Dispersal
mod.poldiso <- lmer(species_CV ~ Pollination * Dispersal + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA" & Dispersal != "WA"))
Anova(mod.poldiso, type = 2)

### Fit model with Pollination x LifeHistory
## Full model removing WA 
mod.pol.lhist.noWA <- lmer(species_CV ~ Pollination * LifeHistory + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA"))

Anova(mod.pol.lhist.noWA, type = 2)

plot_model(mod.pol.lhist.noWA, type = "pred", terms = c("Pollination", "LifeHistory"))

```

Try models that include species_mu
```{r, eval=F}
### Separate model for each predictor ###
mod.pol.mu <- lmer(species_CV ~ Pollination * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Pollination != "WA"))
mod.disp.mu <- lmer(species_CV ~ Dispersal * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2012 %>% filter(Dispersal != "WA"))
mod.leg.mu <- lmer(species_CV ~ Legume * species_mu + X + Y + (1|kartez), na.action = na.omit, data = data.2012)

Anova(mod.pol.mu, type = 2)
Anova(mod.disp.mu, type = 2)
Anova(mod.leg.mu, type = 2)

# Plot marginal effect of each predictor
Pol.mu.plot <- plot_model(mod.pol.mu, type = "pred", terms = "Pollination") +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Pollination Strategy") +
  ylim(c(1.6, 2.8))

Pol.mu.plot

# Look at CV vs. mu for wind vs. animal pollinated species
ggplot(data.2012, aes(x = species_mu, y = species_CV, color = Pollination.new)) +
  geom_point()  +
  geom_smooth(method = "lm", se = TRUE)

#


Disp.mu.plot <- plot_model(mod.disp.mu, type = "pred", terms = "Dispersal")  +
  scale_x_discrete(limits = c("A", "W"),
                   labels = c("Animal", "Wind")) +
  labs(y = "CV of Species Biomass", x = "Dispersal Strategy") +
  ylim(c(1.6, 2.8))

Disp.mu.plot



ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Pol.plot.jpg", Pol.plot, dpi = 600, width = 2000, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Disp.plot.jpg", Disp.plot, dpi = 600, width = 2000, height = 2500, units = "px")

ggsave("~/GitHub/SEV_Species_Interactions_Stability/reports/Leg.plot.jpg", Leg.plot, dpi = 600, width = 2000, height = 2500, units = "px")
```
